Conflict Analysis for Mixed-integer SDPs

### 2022-06-11 tex/ConflictAnalysis.tex
extend description

We consider mixed-integer SDPs of the following form:
\begin{equation}\label{MISDP}
  \begin{aligned}
    \inf \quad & b\T y \\
    \text{s.t.} \quad & \sum_{j=1}^m A^j\, y_j - A^0 \succeq 0, \\
    & D y \geq d,\\
    & \ell \leq y \leq u,\\
    & y_i \in \Z,\quad i \in I.
  \end{aligned}
\end{equation}
In each node, we solve the following relaxations with local bounds $\ell$
and $u$:
\begin{equation}\label{SDP-D}
  \begin{aligned}
    \inf \quad & b\T y \\
    \text{s.t.} \quad & \sum_{j=1}^m A^j\, y_j - A^0 \succeq 0, \\
    & D y \geq d,\\
    & \ell \leq y \leq u,
  \end{aligned}
\end{equation}
where for simplicity of exposition, we assume that all entries of $\ell$
and $u$ are finite.

The general idea of conflict analysis is to generate globally valid
constraints as follows. Let any $\hat{X} \succeq 0$ and $\hat{z} \geq 0$ be
given. Then because $A(y) \define \sum_{j=1}^m A^j\, y_j - A^0$ is positive
semidefinite for all feasible $y \in \R^m$, we have
$\skal{A(y)}{\hat{X}} \geq 0$.  Similarly, the aggregated inequalities
$\hat{z}\T D \geq \hat{z}\T d$ hold. Together this yields:
\begin{align}
  & \skal{A(y)}{\hat{X}} + \hat{z}\T D \geq \hat{z}\T d\notag\\
  \Leftrightarrow\quad
  & \sum_{j=1}^m \skal{A^j}{\hat{X}} y_j + \hat{z}\T D + \geq
    \skal{A^0}{\hat{X}} + \hat{z}\T d,\label{eq:ConflictCut}
\end{align}
which is valid for all feasible $y$. Clearly,
Inequality~\eqref{eq:ConflictCut} is redundant for continuous $y$, but it
might allow to cut off certain mixed-integer solutions~$y$. In the implementation,
we only propagate bounds using~\eqref{eq:ConflictCut}.

More generally, we can also use the variable bounds $\ell \leq y$ and
$y \leq u$. If $\hat{r}^\ell$, $\hat{r}^u \geq 0$ are the corresponding
multiliers, this yields:
\begin{align}
  & \skal{A(y)}{\hat{X}} + \hat{z}\T D + (\hat{r}^\ell)\T y - (\hat{r}^u)\T y
    \geq \hat{z}\T d + (\hat{r}^\ell)\T \ell + (\hat{r}^u)\T u\notag\\
  \Leftrightarrow
  & \sum_{j=1}^m \skal{A^j}{\hat{X}} y_j + \hat{z}\T D +
    (\hat{r}^\ell)\T y - (\hat{r}^u)\T y \geq \skal{A^0}{\hat{X}} +
    \hat{z}\T d + (\hat{r}^\ell)\T \ell + (\hat{r}^u)\T u,\label{eq:ConflictCut2}
\end{align}
which is again valid for all feasible $y$.

% -------------------------------------------------------------------------
\section{Generation of Conflict Constraints}

Let $X$ be the dual variables for the SDP-constraint $A(y) \succeq 0$, $z$
be the dual variables for $Dy \geq d$, and $r^\ell$ and $r^u$ be the dual
variables for the bound constraints $y \geq \ell$ and $-y \geq -u$,
respectively. Then the corresponding primal problem for~\eqref{SDP-D} is:
\begin{equation}\label{SDP-P}
  \begin{aligned}
     \sup \quad & \skal{A^0}{X} + z\T d + \ell\T r^\ell - u\T r^u\\
     \text{s.t.} \quad & \skal{A^j}{X} + D\T z + r^\ell_j - r^u_j = b_j && \forall \, j \in [m],\\
     & X \succeq 0,\; z, \; r^\ell,\; r^u \geq 0.
  \end{aligned}
\end{equation}
Any feasible solution can be used to construct~\eqref{eq:ConflictCut}
or~\eqref{eq:ConflictCut2}.

Similarly, a primal ray $(\hat{X}, \hat{z}, \hat{r}^\ell, \hat{r}^u)$,
i.e.,
\begin{equation}\label{eq:ray}
\begin{aligned}
  & \skal{A^j}{\hat{X}} + D\T \hat{z} + \hat{r}^\ell_j - \hat{r}^u_j = 0 && \forall\; j \in [m],\\
  & \skal{A^0}{\hat{X}} + d\T \hat{z} + \ell\T \hat{r}^\ell - u\T \hat{r}^u > 0,
\end{aligned}
\end{equation}
yields an inequality as in~\eqref{eq:ConflictCut}
or~\eqref{eq:ConflictCut2}.

If $\ell \leq u$ and~\eqref{eq:ray} has a feasible solution
$(\tilde{X}, \tilde{r}^\ell, \tilde{r}^u)$, then
$(\hat{X}, \hat{r}^\ell, \hat{r}^u)$ with $\hat{X} = \tilde{X}$ and
$\hat{r}^\ell_j = - \min\,\{\skal{A^j}{\tilde{X}}, 0\} \geq 0$ and
$\hat{r}^u_j = \max\,\{\skal{A^j}{\tilde{X}}, 0\} \geq 0$ is feasible
for~\eqref{eq:ray}. Indeed, it is clear that the equations
of~\eqref{eq:ray} are satisfied. Therefore consider the contribution to
$\ell\T \hat{r}^\ell - u\T \hat{r}^u$ in component $j$:
\begin{align*}
  \ell_j \hat{r}^\ell_j - u_j \hat{r}^u_j & = - \ell_j\, \min\,\{\skal{A^j}{\tilde{X}}, 0\} - u_j\, \max\,\{\skal{A^j}{\tilde{X}}, 0\}\\
  & = - \ell_j\, \min\,\{\tilde{r}^u_j - \tilde{r}^\ell_j, 0\} - u_j\, \max\,\{\tilde{r}^u_j - \tilde{r}^\ell_j, 0\}\\
  & = \begin{cases}
    - \ell_j\, (\tilde{r}^u_j - \tilde{r}^\ell_j) & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j < 0,\\
    - u_j\, (\tilde{r}^u_j - \tilde{r}^\ell_j) & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j \geq 0,
  \end{cases}\\
  & = \begin{cases}
    - \ell_j\, \tilde{r}^u_j + \ell_j \tilde{r}^\ell_j & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j < 0,\\
    - u_j\, \tilde{r}^u_j + u_j \tilde{r}^\ell_j & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j \geq 0,
  \end{cases}\\
  & \geq \begin{cases}
    - u_j\, \tilde{r}^u_j + \ell_j \tilde{r}^\ell_j & \text{if }  \tilde{r}^u_j - \tilde{r}^\ell_j < 0,\\
    - u_j\, \tilde{r}^u_j + \ell_j \tilde{r}^\ell_j & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j \geq 0,
  \end{cases}\\
  & = \ell_j \tilde{r}^\ell_j - u_j \tilde{r}^u_j.
\end{align*}
Therefore the inequality in~\eqref{eq:ray} holds as well.

This implies that
\begin{align*}
  0 & < \skal{A^0}{\hat{X}} + \ell\T \hat{r}^\ell - u\T \hat{r}^u\\
  & = \skal{A^0}{\hat{X}} + \sum_{j=1}^m \ell_j \hat{r}^\ell_j -
    \sum_{j=1}^m u_j \hat{r}^u_j\\
  & = \skal{A^0}{\hat{X}} - \sum_{j: \skal{A^j}{\hat{X}} < 0} \ell_j \skal{A^j}{\hat{X}} -
    \sum_{j: \skal{A^j}{\hat{X}} > 0} u_j \skal{A^j}{\hat{X}}.
\end{align*}
This implies
\begin{align*}
  \sum_{j: \skal{A^j}{\hat{X}} < 0} \ell_j \skal{A^j}{\hat{X}} +
    \sum_{j: \skal{A^j}{\hat{X}} > 0} u_j \skal{A^j}{\hat{X}} < \skal{A^0}{\hat{X}}.
\end{align*}
However, any feasible solution $\hat{y}$ satisfies $A(\hat{y}) \succeq 0$ and $\ell
\leq \hat{y} \leq u$ and therefore
\begin{align*}
  0 & \leq \skal{A(\hat{y})}{\hat{X}} =
      \sum_{j=1}^m \skal{A^j}{\hat{X}}\, \hat{y}_j + \sum_{j=1}^m
      \skal{A^j}{\hat{X}}\, \hat{y}_j - \skal{A^0}{\hat{X}}\\
    & \leq \sum_{j: \skal{A^j}{\hat{X}} < 0} \skal{A^j}{\hat{X}}\, \ell_j +
    \sum_{j: \skal{A^j}{\hat{X}} > 0} \skal{A^j}{\hat{X}} u_j -
      \skal{A^0}{\hat{X}},
\end{align*}
a contradition. This implies that the valid linear inequality
$\skal{A(\hat{y})}{\hat{X}} \geq 0$ shows infeasibility with respect to the
local bounds.

% -------------------------------------------------------------------------
\section{Infeasibility Proof Analysis}

>>>>

We consider mixed-integer SDPs of the following form:
\begin{equation}\label{MISDP}
  \begin{aligned}
    \inf \quad & b\T y \\
    \text{s.t.} \quad & \sum_{j=1}^m A^j\, y_j - A^0 \succeq 0, \\
    & D y \geq d,\\
    & \ell \leq y \leq u,\\
    & y_i \in \Z,\quad i \in I.
  \end{aligned}
\end{equation}
In each node, we solve the following relaxations with local bounds $\ell$
and $u$:
\begin{equation}\label{SDP-D}
  \begin{aligned}
    \inf \quad & b\T y \\
    \text{s.t.} \quad & \sum_{j=1}^m A^j\, y_j - A^0 \succeq 0, \\
    & D y \geq d,\\
    & \ell \leq y \leq u,
  \end{aligned}
\end{equation}
where for simplicity of exposition, we assume that all entries of $\ell$
and $u$ are finite. Moreover, we always assume that $\ell \leq u$ holds.

The general idea of conflict analysis is to generate globally valid
constraints as follows. Let any $\hat{X} \succeq 0$ and $\hat{z} \geq 0$ be
given. Then because $A(y) \define \sum_{j=1}^m A^j\, y_j - A^0$ is positive
semidefinite for all feasible $y \in \R^m$, we have
$\skal{A(y)}{\hat{X}} \geq 0$.  Similarly, the aggregated inequalities
$\hat{z}\T D y \geq \hat{z}\T d$ hold. Together this yields:
\begin{align}
  & \skal{A(y)}{\hat{X}} + \hat{z}\T D y \geq \hat{z}\T d\notag\\
  \Leftrightarrow\quad
  & \sum_{j=1}^m \skal{A^j}{\hat{X}} y_j + \hat{z}\T D y \geq
    \skal{A^0}{\hat{X}} + \hat{z}\T d,\label{eq:ConflictCut}
\end{align}
which is valid for all feasible $y$. Clearly,
Inequality~\eqref{eq:ConflictCut} is redundant for continuous $y$, but it
might allow to cut off certain mixed-integer solutions~$y$. In the implementation,
we only propagate bounds using~\eqref{eq:ConflictCut}.

More generally, we can also use the variable bounds $\ell \leq y$ and
$y \leq u$. If $\hat{r}^\ell$, $\hat{r}^u \geq 0$ are the corresponding
multiliers, this yields:
\begin{align}
  & \skal{A(y)}{\hat{X}} + \hat{z}\T D y + (\hat{r}^\ell)\T y - (\hat{r}^u)\T y
    \geq \hat{z}\T d + (\hat{r}^\ell)\T \ell + (\hat{r}^u)\T u\notag\\
  \Leftrightarrow\quad
  & \sum_{j=1}^m \skal{A^j}{\hat{X}} y_j + \hat{z}\T D y +
    (\hat{r}^\ell)\T y - (\hat{r}^u)\T y \geq \\
  & \qquad \skal{A^0}{\hat{X}} +
    \hat{z}\T d + (\hat{r}^\ell)\T \ell + (\hat{r}^u)\T u,\label{eq:ConflictCut2}
\end{align}
which is again valid for all feasible $y$.

It seems that~\eqref{eq:ConflictCut2} are less helpful
than~\eqref{eq:ConflictCut}. However, one can cancel some coefficients
in~\eqref{eq:ConflictCut} using variable bounds -- see
Section~\ref{sec:Canceling}.

% -------------------------------------------------------------------------
% -------------------------------------------------------------------------
\section{Generation of Conflict Constraints}

% -------------------------------------------------------------------------
\subsection{Feasible Problems}

Let $X$ be the dual variables for the SDP-constraint $A(y) \succeq 0$, $z$
be the dual variables for $Dy \geq d$, and $r^\ell$ and $r^u$ be the dual
variables for the bound constraints $y \geq \ell$ and $-y \geq -u$,
respectively. Then the corresponding primal problem for~\eqref{SDP-D} is:
\begin{equation}\label{SDP-P}
  \begin{aligned}
     \sup \quad & \skal{A^0}{X} + z\T d + \ell\T r^\ell - u\T r^u\\
     \text{s.t.} \quad & \skal{A^j}{X} + (D\T z)_j + r^\ell_j - r^u_j = b_j && \forall \, j \in [m],\\
     & X \succeq 0,\; z, \; r^\ell,\; r^u \geq 0.
  \end{aligned}
\end{equation}
Any feasible solution $(\hat{X}, \hat{z}, \hat{r}^\ell, \hat{r}^u)$ can be
used to construct~\eqref{eq:ConflictCut} or~\eqref{eq:ConflictCut2}.

Moreover, one can take an objective cutoff into account. Namely, assume
that $y^\star$ is a feasible solution to~\eqref{MISDP}. We are then only
interested in solutions that are at least as good, i.e.,
$b\T y \leq b\T y^\star$ or $- b\T y \geq - b\T y^\star$. Adding this
inequality to~\eqref{eq:ConflictCut} yields:
\begin{align*}
  & \sum_{j=1}^m \skal{A^j}{\hat{X}} y_j + \hat{z}\T D y - b\T y \geq
    \skal{A^0}{\hat{X}} + \hat{z}\T d - b\T y^\star\\
  \Leftrightarrow\quad
  & \sum_{j=1}^m \big(\skal{A^j}{\hat{X}} + (\hat{z}\T D)_j - b_j\big) y_j \geq
    \skal{A^0}{\hat{X}} + \hat{z}\T d - b\T y^\star.
\end{align*}

% -------------------------------------------------------------------------
\subsection{Infeasible Problems}

Similarly, a primal ray $(\hat{X}, \hat{z}, \hat{r}^\ell, \hat{r}^u)$,
i.e., it satsifies the system
\begin{equation}\label{eq:ray}
\begin{aligned}
  & \skal{A^j}{X} + (D\T z)_j + r^\ell_j - r^u_j = 0 && \forall\; j \in [m],\\
  & \skal{A^0}{X} + d\T z + \ell\T r^\ell - u\T r^u > 0,
\end{aligned}
\end{equation}
yields an inequality as in~\eqref{eq:ConflictCut}
or~\eqref{eq:ConflictCut2}.

In this case, one can prove that~\eqref{eq:ConflictCut} actually proves
infeasibility for the local bounds~$\ell$ and $u$. To this end, we first
show the following:

\begin{lemma}\label{lemma:ComplemenatrySolution}
  If~\eqref{eq:ray} has a feasible solution, then there exists a feasible
  solution in which $r_j^\ell \cdot r_j^u = 0$ for all $j \in [m]$, i.e.,
  at most one of $r_j^\ell$ and $r_j^u$ is nonzero.
\end{lemma}

\begin{proof}
  Let $(\hat{X}, \hat{z}, \tilde{r}^\ell, \tilde{r}^u)$ be a feasible
  solution of~\eqref{eq:ray}. Then define
  \[
    \hat{r}^\ell_j \define - \min\,\{\skal{A^j}{\hat{X}} + (D\T \hat{z})_j,\; 0\} \geq 0,
    \quad
    \hat{r}^u_j = \max\,\{\skal{A^j}{\hat{X}} + (D\T \hat{z})_j,\; 0\} \geq 0.
  \]
  We claim that $(\hat{X}, \hat{z}, \hat{r}^\ell, \hat{r}^u)$ is feasible
  for~\eqref{eq:ray}. Note that by construction we have
  $\hat{r}^\ell_j \cdot \hat{r}^u_j = 0$ for all $j \in [m]$.

  First observere that the equations of~\eqref{eq:ray} are
  satisfied. Therefore consider the contribution to
  $\ell\T \hat{r}^\ell - u\T \hat{r}^u$ in component $j$:
  \begin{align*}
    \ell_j \hat{r}^\ell_j - u_j \hat{r}^u_j & = - \ell_j\, \min\,\{\skal{A^j}{\hat{X}} + (D\T \hat{z})_j, 0\} - u_j\, \max\,\{\skal{A^j}{\hat{X}} + (D\T \hat{z})_j, 0\}\\
    & = - \ell_j\, \min\,\{\tilde{r}^u_j - \tilde{r}^\ell_j, 0\} - u_j\, \max\,\{\tilde{r}^u_j - \tilde{r}^\ell_j, 0\}\\
    & = \begin{cases}
      - \ell_j\, (\tilde{r}^u_j - \tilde{r}^\ell_j) & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j < 0,\\
      - u_j\, (\tilde{r}^u_j - \tilde{r}^\ell_j) & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j \geq 0,
    \end{cases}\\
    & = \begin{cases}
      - \ell_j\, \tilde{r}^u_j + \ell_j \tilde{r}^\ell_j & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j < 0,\\
      - u_j\, \tilde{r}^u_j + u_j \tilde{r}^\ell_j & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j \geq 0,
    \end{cases}\\
    & \geq \begin{cases}
      - u_j\, \tilde{r}^u_j + \ell_j \tilde{r}^\ell_j & \text{if }  \tilde{r}^u_j - \tilde{r}^\ell_j < 0,\\
      - u_j\, \tilde{r}^u_j + \ell_j \tilde{r}^\ell_j & \text{if } \tilde{r}^u_j - \tilde{r}^\ell_j \geq 0,
    \end{cases}\\
    & = \ell_j \tilde{r}^\ell_j - u_j \tilde{r}^u_j.
  \end{align*}
  Therefore the inequality in~\eqref{eq:ray} holds as well.
\end{proof}

\begin{proposition}
  Let $(\hat{X}, \hat{z})$ be part of a solution of~\eqref{eq:ray}. Then~\eqref{eq:ConflictCut}, i.e.,
  \[
    \skal{A(y)}{\hat{X}} + \hat{z}\T D \geq \hat{z}\T d
  \]
  is infeasible with respect to the local bounds $\ell$ and $u$.
\end{proposition}

\begin{proof}
  One the one hand, any feasible solution $\hat{y}$ of~\eqref{MISDP} satisfies
  $A(\hat{y}) \succeq 0$, $D \hat{y} \geq d$, and
  $\ell \leq \hat{y} \leq u$. Defining
  $\hat{s}_j \define \skal{A^j}{\hat{X}} + (\hat{z}\T D)_j$ for
  $j \in [m]$, we get from~\eqref{eq:ConflictCut}:
  \begin{align*}
    \skal{A^0}{\hat{X}} + \hat{z}\T d & \leq \sum_{j=1}^m \big(\skal{A^j}{\hat{X}} + (\hat{z}\T D)_j\big)\, \hat{y}_j
    = \sum_{j=1}^m \hat{s}_j\, \hat{y}_j\\
    & \leq \sum_{j: \hat{s}_j < 0} \hat{s}_j\, \ell_j +
        \sum_{j: \hat{s}_j > 0} \hat{s}_j\, u_j.
  \end{align*}

  On the other hand let $(\hat{X}, \hat{z}, \hat{r}^\ell, \hat{r}^u)$ be a
  feasible solution of~\eqref{eq:ray} that satisfies the conditions of
  Lemma~\ref{lemma:ComplemenatrySolution}. Then
  $\hat{r}^u - \hat{r}^\ell = \hat{s}$.  We thus obtain
  \begin{align*}
    0 & < \skal{A^0}{\hat{X}} + d\T \hat{z} + \ell\T \hat{r}^\ell - u\T \hat{r}^u\\
      & = \skal{A^0}{\hat{X}} + d\T \hat{z} + \sum_{j=1}^m \ell_j \hat{r}^\ell_j -
        \sum_{j=1}^m u_j \hat{r}^u_j\\
      & = \skal{A^0}{\hat{X}}  + d\T \hat{z} - \sum_{j: \hat{s}_j < 0} \ell_j\, \hat{s}_j -
        \sum_{j: \hat{s}_j > 0} u_j\, \hat{s}_j,
  \end{align*}
  a contradition.
\end{proof}

% -------------------------------------------------------------------------
\subsection{Canceling}
\label{sec:Canceling}

Starting from~\eqref{eq:ConflictCut}, one can possibly cancel coefficients
as follows. Again using
$s_j \define \skal{A^j}{\hat{X}} + (\hat{z}\T D)_j$, we get
from~\eqref{eq:ConflictCut}:
\begin{equation}\label{eq:ConflictCut3}
  \sum_{j=1}^m s_j\, y_j \geq \skal{A^0}{\hat{X}} + \hat{z}\T d.
\end{equation}

Assume that $s_j < 0$ for some $j \in [m]$ and the global lower bound of
$y_j$ is at least 0. Then we can add the valid inequality $-s_j\, y_j \geq 0$
to~\eqref{eq:ConflictCut3}, which eliminates variable~$y_j$ from the
inequality. Similarly, if $s_j > 0$ for some $j \in [m]$ and the global
upper bound of $y_j$ is at most 0, we can add $s_j\, y_j \leq 0$ to
eliminate variable~$y_j$.

Witzig et al.~\cite{WitBH17} (see also Witzig~\cite{Wit22}) applied
cancelation to continuous variables, but restricts attention to variables
that are at their global bounds at the node at which the constraint is
generated.

\begin{small}
   \bibliographystyle{plain}
   \bibliography{ConflictAnalysis.bib}
\end{small}
